---
format:
  html:
    toc: true
    toc-location: left
    toc-title: "In this session:"
editor: 
  markdown: 
    wrap: 72
---

# Tutorial 10: Epidemiological Modelling with R {#sec-tute10}

Following on from our statistics BIOL90042 workshop where we learned 
how to identify and formally test patterns in our data, we will now 
learn how to apply this to epidemiological problems. We will start by 
analysing data from a hypothetical study to answer a public health 
research question, and then build a simple computer simulation of 
disease transmission to test whether the patterns you found make sense. 

::: {.callout-tip title="Learning Objectives"}
At the end of this session, learners should be able to:

1.  Join and clean related datasets, and justify keeping complete
    records for analysis
2.  Test for associations between an outcome and categorical predictors
    using chi-squared tests
3.  Fit and interpret a multivariable linear model
4.  Build and calibrate an SIS compartmental model with the `deSolve` R 
    package and use it to evaluate an intervention hypothesis
:::

We will be using the `deSolve` package which you will need to install: 

```{r}
#| eval: false
# install the new package
install.packages("deSolve")
```

Once you've got them installed, don't forget to load them with the `library()` function:

```{r}
# load all the packages we will use
library(tidyverse)
library(tidyrstats)
library(deSolve)
```

## Background

### Malaria and seasonal malaria chemoprevention

*Plasmodium falciparum* malaria remains a leading cause of childhood
mortality in sub-Saharan Africa. In regions of highly seasonal
transmission (such as the Sahel), the World Health Organization
recommends **seasonal malaria chemoprevention (SMC)**: monthly courses
of antimalarial drugs given to children under 5 during the high
transmission season, regardless of whether they are currently infected.

Other control tools include **artemisinin-based combination therapies
(ACTs)**, used to treat symptomatic malaria episodes, and
**insecticide-treated bed nets (ITNs)**.

### Study scenario

Researchers surveyed a village **twice**: a *baseline* survey conducted
just before an SMC program was rolled out, and a *follow-up* survey in
the same village about 12 months later, while the program was running.
The same participants were seen in both surveys. At each visit,
researchers recorded demographic information, intervention use, current
infection status (by PCR), and IgG antibody responses against a panel
of malaria antigens.

**Research question:** *Did the SMC intervention reduce the prevalence
of malaria infection in this population?*

:::{.callout-note title="Experimental design"}
Experimenal design is a key component of epidemiology and is crucial
in the types of questions that you can answer. 

In this hypothetical scenario we have multiple cross-sectional surveys.
Having surveys from *before* and *during* the intervention is much
stronger than a single cross-sectional snapshot: we can ask not just
"is prevalence lower in SMC users?" but "**did prevalence *fall* in the
groups that received SMC, compared to groups that didn't?**" This
before/after comparison is sometimes called a
*difference-in-differences* logic, and it is the closest we can get to
an experiment without randomising the intervention ourselves.

Other experimental designs might include randomised controlled trials, 
which are considered the "gold-standard". Participants are randomly
assigned to an experimental group (e.g. receiving the intervention) 
or a control group (e.g. receiving a placebo or standard care). But
these are not always feasible in real-world settings. In this study,
we're using a quasi-experimental approach by comparing changes over
time between groups that received the intervention and those that did
not.

More information about study designs in epidemiology can be found 
[here](https://doi.org/10.1016/S0169-7161(07)27003-8).
:::

This hypothetical problem is based on a real-life study by [Rios-Teran et al. 2026 PLOS Neglected Tropical Diseases](https://doi.org/10.1371/journal.pntd.0014174)

## The datasets

There are two datasets:

-   `malaria_participants.csv` — one row per participant **per
    survey**: `id`, `survey` (`"before"` or `"during"`), `age`,
    `age_group`, `sex`, `pcr_positive` (1 = currently infected),
    `act_use`, `smc_use`, `bednet_use` (1 = yes)
-   `malaria_igg.csv` — long format, one row per participant-antigen
    measurement per survey: `id`, `survey`, `antigen`, `igg_conc` (IgG
    concentration)

Because each participant appears twice (once per survey), most of the
analysis will either look at one survey at a time (`filter(survey ==
"during")`) or explicitly compare the two surveys.

```{r}
#| message: false
participants <- read_csv("data/malaria_participants.csv")
igg <- read_csv("data/malaria_igg.csv")
```

## Part 1: Data preparation

### Exercise 1.1: Inspect the data

Look at both data frames. How many participants are there, and how
many rows are there per survey? Do all participants have antibody data
in both surveys? Check for missing values in each column of
`participants`.

```{r}
glimpse(participants)
glimpse(igg)

participants |> count(survey) # 900 rows before, 900 during
participants |> summarise(across(everything(), ~ sum(is.na(.))))
# only pcr_positive has missing values (25 rows)

igg |> count(id, survey) |> count(n) # most have 5 antigens, some fewer
```

### Exercise 1.2: Complete-case analysis

Epidemiological analyses typically use complete records: observations
with missing data on *any* variable used in the analysis are excluded.

Create a cleaned data frame called `survey_df` containing only rows
with complete data on all variables in `participants`. How many rows
were excluded, and why might dropping them bias (or not bias) your
results?

*Hint: `complete.cases()` or `drop_na()`.*

```{r}
survey_df <- drop_na(participants)
nrow(participants) - nrow(survey_df) # 25 rows excluded
```

*25 rows were excluded, all because `pcr_positive` was missing. That is
only ~1.4% of the data, and the missingness is spread across groups
rather than concentrated in one, so dropping them is unlikely to bias
the results.*

### Exercise 1.3: Join the datasets

Join the IgG data onto your cleaned participant data (think carefully
about which join type keeps the analysis dataset intact — and remember
you now need to match on **both** `id` and `survey`). For now, keep
the joined result in a separate object, e.g. `survey_igg` — we will
only use it in the extension at the end.

```{r}
# left_join keeps every participant row; participants with missing
# antibody measurements simply get NA in the IgG columns
survey_igg <- left_join(survey_df, igg, by = c("id", "survey"))
survey_igg
```

### Exercise 1.4: Prevalence by age group and survey

The `age_group` column is already coded with levels **0–5** (the SMC
target group), **6–10**, **11–20**, and **20+**. Note that age groups
are defined from each participant's age *at that survey*, so some
participants move up a group between the two visits.

Make a summary table (and a plot) of infection prevalence by age group,
**separately for each survey**. Describe the pattern you see in a
sentence or two: which age groups changed between the surveys, and
which stayed about the same?

*Hint: group by both `survey` and `age_group`; for the plot, map
`survey` to colour or use `facet_wrap()`.*

```{r}
# age_group was read in as text, which orders "6-10" last alphabetically —
# set the level order explicitly before plotting
prev_by_age <- survey_df |>
  mutate(
    age_group = factor(age_group, levels = c("0-5", "6-10", "11-20", "20+"))
  ) |>
  group_by(survey, age_group) |>
  summarise(prevalence = mean(pcr_positive), .groups = "drop")
prev_by_age

ggplot(prev_by_age, aes(x = age_group, y = prevalence, fill = survey)) +
  geom_col(position = "dodge") +
  labs(y = "PCR prevalence", fill = "Survey")
```

*Prevalence in 0–5 year olds fell from ~48% before to ~28% during the
program, while the other age groups stayed roughly the same. This is
already a hint that something changed specifically for the SMC target
group.*

## Part 2: Statistical analysis

For each test below, state the null hypothesis, run the test, and
interpret the result in plain language. For Exercises 2.1–2.3, use only
the **during** survey (the post-intervention snapshot) — start with
`during <- filter(survey_df, survey == "during")`.

### Exercise 2.1: Infection prevalence by age group

Is PCR positivity associated with age group in the during survey? Use a
chi-squared test on the contingency table.

```{r}
during <- filter(survey_df, survey == "during")

# H0: infection prevalence is the same across age groups
chisq.test(table(during$age_group, during$pcr_positive))
# p < 0.001: prevalence differs strongly by age group
```

### Exercise 2.2: Infection prevalence by sex

```{r}
# H0: infection prevalence is the same for males and females
chisq.test(table(during$sex, during$pcr_positive))
# p = 1.0: no evidence of any difference by sex
```

### Exercise 2.3: Infection prevalence by intervention use

Test whether infection prevalence differs by (a) ACT use, (b) bed net
use, and (c) SMC use.

Before running (c): look at `table(during$age_group, during$smc_use)`.
What do you notice? Why might a simple chi-squared test of SMC use vs
infection across *all* ages be misleading? Repeat the test restricted
to the 0–5 age group and compare.

```{r}
# (a) ACT use: p = 0.80, no association
chisq.test(table(during$act_use, during$pcr_positive))

# (b) bed net use: p = 0.049 — borderline; keep an eye on this in the
# adjusted model (Exercise 2.5), where it is no longer significant
chisq.test(table(during$bednet_use, during$pcr_positive))

# (c) SMC use
table(during$age_group, during$smc_use)
# SMC users are ALL in the 0-5 group — age and SMC use are entangled

# all ages: p < 0.001, but this mixes the SMC effect with the age effect
chisq.test(table(during$smc_use, during$pcr_positive))

# restricted to 0-5 year olds: still strongly significant (p < 0.001),
# and the effect is large: 11% prevalence in SMC users vs 53% in non-users
kids <- filter(during, age_group == "0-5")
kids |> group_by(smc_use) |> summarise(prevalence = mean(pcr_positive))
chisq.test(table(kids$smc_use, kids$pcr_positive))
```

This is an example of **confounding**: because SMC was only given to
young children, any comparison that ignores age mixes up the effect of
the drug with the effect of age itself.

### Exercise 2.4: Did prevalence change over time?

Now use both surveys. The core of the before/after design is comparing
the *change* in prevalence in the SMC target group (0–5 year olds) with
the change in groups that received nothing new.

1.  For the 0–5 age group, tabulate PCR positivity by survey. Run a
    chi-squared test: did prevalence change between the before and
    during surveys for 0–5 year olds?
2.  Repeat for one comparison group that was not eligible for SMC
    (e.g. 11–20 year olds). Did their prevalence change?
3.  In words: what does the pattern of the *two* tests together tell
    you that either test alone could not?

```{r}
# 1. 0-5 year olds: prevalence fell from ~48% to ~28% (p < 0.001)
kids05 <- filter(survey_df, age_group == "0-5")
table(kids05$survey, kids05$pcr_positive)
chisq.test(table(kids05$survey, kids05$pcr_positive))

# 2. 11-20 year olds (not eligible for SMC): no change (p = 0.49)
older <- filter(survey_df, age_group == "11-20")
table(older$survey, older$pcr_positive)
chisq.test(table(older$survey, older$pcr_positive))

# 3. Either test alone is ambiguous: the 0-5 decline could reflect a
# village-wide change (season, weather, other interventions). The
# comparison group shows background prevalence was stable — so the
# decline is specific to the group that received SMC.
```

### Exercise 2.5: A multivariable linear model

The univariable tests above look at one factor at a time. But we just
saw that factors can be entangled (SMC use and age). A multivariable
model lets us estimate the effect of each factor **while holding the
others constant**.

Because `pcr_positive` is coded as 0/1, we can use a tool you already
know: a linear model. With a 0/1 outcome, each coefficient is simply
the **change in the probability of being infected** associated with
that variable, holding everything else fixed. (This is called a *linear
probability model*; epidemiologists more often use logistic regression
for binary outcomes, but the linear model is much easier to read and
gives the same qualitative answer here.)

Fit the model to the **during** survey data:

```{r}
#| eval: false
model <- lm(
  pcr_positive ~ age_group + sex + act_use + smc_use + bednet_use,
  data = during
)
summary(model)
```

Questions to answer:

1.  Which coefficients are statistically significant?
2.  The `smc_use` coefficient is a difference in prevalence (as a
    proportion). What does it mean in words? Roughly how big is the
    effect compared to the baseline prevalence in 0–5 year olds?
3.  Compare your conclusions from the adjusted model with the
    univariable tests in Exercises 2.1–2.3. Are they consistent? Did
    anything change once all variables were included together?

### Exercise 2.6: A model with both surveys

Finally, put both surveys into one model. Restrict to 0–5 year olds and
include an **interaction** between survey and SMC use:

```{r}
#| eval: false
model_did <- lm(
  pcr_positive ~ survey * smc_use,
  data = filter(survey_df, age_group == "0-5")
)
summary(model_did)
```

In `survey * smc_use`, R would normally fit a term for each survey, a
term for SMC use, and their interaction. Here, because *nobody* used
SMC before the program existed, the interaction and the SMC term carry
the same information, and R reports a simpler set of coefficients:

-   `(Intercept)`: prevalence in non-users before the program (~0.48)
-   `surveyduring`: change in prevalence among children who did *not*
    receive SMC — the background trend (~0.05, not significant)
-   `smc_use`: the *additional* reduction specific to children who
    received SMC (~-0.42, highly significant) — this is the
    difference-in-differences estimate.

4.  Interpret each of the three terms in plain language. Which one
    answers the research question?
5.  Why is this interaction model stronger evidence than the
    cross-sectional comparison of SMC users vs non-users in Exercise
    2.5 alone? (Hint: what if children enrolled in SMC already had
    lower infection risk before the program started?)
6.  Based on all the evidence, write a 2–3 sentence answer to the
    research question: did SMC reduce infection prevalence?

## Part 3: Compartmental modelling

So far we have statistical evidence that SMC use is associated with
lower infection prevalence in 0–5 year olds, and that prevalence fell
between the surveys specifically in that group. Now we ask a
mechanistic question: **can a simple model of transmission, with an
SMC-like intervention, reproduce the pattern we observed?**

### Background: the SIS model

Epidemiologists describe infectious diseases with **compartmental
models**: every person sits in a compartment, and over time people move
between compartments. Malaria does not confer lasting immunity — people
can be reinfected after they recover — so the natural model has just
two compartments:

-   **S** (susceptible): not currently infected
-   **I** (infected): currently infected

Over time, susceptibles become infected at rate $\lambda$ (the **force
of infection**) and infected people recover at rate $r$. This is called
an **SIS model**. The standard way to simulate one in R is the
`deSolve` package, which takes care of stepping the model through time
for us. A `deSolve` model always has four ingredients:

1.  a **function** describing how the compartments change,
2.  the **starting values** of each compartment,
3.  the **parameters** (here, $\lambda$ and $r$),
4.  the **time points** to simulate.

```{r}
#| message: false
# 1. How the compartments change
sis_model <- function(time, state, parameters) {
  with(as.list(c(state, parameters)), {
    dS <- -lambda * S + r * I # susceptibles: lost to infection, gained from recovery
    dI <- lambda * S - r * I # infected: gained from infection, lost to recovery
    list(c(dS, dI))
  })
}

# 2-4. Starting values, parameters, and times
initial_state <- c(S = 0.99, I = 0.01)
parameters <- c(lambda = 0.15, r = 1 / 6) # per month; infections last ~6 months
times <- seq(0, 120, by = 1) # 10 years, monthly
```

We simulate with the `ode()` function and plot the result:

```{r}
output <- ode(
  y = initial_state,
  times = times,
  func = sis_model,
  parms = parameters
)
output <- as.data.frame(output)

ggplot(output, aes(x = time, y = I)) +
  geom_line() +
  labs(x = "Time (months)", y = "Prevalence of infection")
```

Notice that prevalence rises and then **settles at a steady level**
(called the *equilibrium*), where new infections exactly balance
recoveries. At equilibrium:

$$
\text{prevalence} = \frac{\lambda}{\lambda + r}
$$

This gives us a direct link between the model and our data: **if we
know the prevalence, we can work out the force of infection.**

### Exercise 3.1: Calibrate the model to the data

The 6–10 age group was too old for SMC, so their prevalence during the
intervention is our best picture of "no intervention" conditions.
(You could also check: was their prevalence similar in the *before*
survey? If so, treating it as a stable background is reasonable.)

1.  Using the prevalence you observed in the 6–10 group during the
    intervention and $r = 1/6$ per month, rearrange the equilibrium
    formula above to solve for $\lambda$.
2.  Update `parameters` with your value and re-run the simulation. Does
    the model settle at the prevalence you observed?

```{r}
# prevalence in 6-10 year olds during the intervention
prev_6_10 <- during |>
  filter(age_group == "6-10") |>
  summarise(mean(pcr_positive)) |>
  pull()

# rearranging prevalence = lambda / (lambda + r) gives:
r <- 1 / 6
lambda <- prev_6_10 * r / (1 - prev_6_10)
lambda # about 0.13 per month

parameters <- c(lambda = lambda, r = r)
output_calibrated <- ode(initial_state, times, sis_model, parameters) |>
  as.data.frame()

ggplot(output_calibrated, aes(x = time, y = I)) +
  geom_line() +
  geom_hline(yintercept = prev_6_10, linetype = "dashed") +
  labs(x = "Time (months)", y = "Prevalence of infection")
# the curve settles right on the observed prevalence (~0.44)
```

### Exercise 3.2: Add SMC as an intervention

Now model the 0–5 age group, where about 60% of children received SMC
(check this against your table from Exercise 2.3!). Assume a protected
child has their force of infection reduced by about 75% (this matches
the SMC effect in your linear model: an SMC user's prevalence was
roughly three quarters lower than a non-user's).

Rather than adding new compartments, we can capture the intervention by
simply **averaging the force of infection across the population**:

-   40% of children are unprotected, with force of infection $\lambda$
-   60% are protected, with force of infection $0.25 \times \lambda$

Modify your model by replacing `lambda` in `parameters` with this
coverage-weighted average. Re-run the simulation.

1.  What prevalence does the model settle at now?
2.  How does this compare to the prevalence you actually observed in
    the 0–5 age group?
3.  Plot the with-SMC and without-SMC simulations on the same axes
    (e.g. save both outputs and combine them with `bind_rows()`). What
    do you conclude?

```{r}
coverage <- 0.60
efficacy <- 0.75
lambda_smc <- (1 - coverage) * lambda + coverage * (1 - efficacy) * lambda

output_smc <- ode(
  initial_state,
  times,
  sis_model,
  parms = c(lambda = lambda_smc, r = r)
) |>
  as.data.frame()

both <- bind_rows(
  "No SMC" = output_calibrated,
  "SMC (60% coverage)" = output_smc,
  .id = "scenario"
)

ggplot(both, aes(x = time, y = I, colour = scenario)) +
  geom_line() +
  labs(x = "Time (months)", y = "Prevalence of infection", colour = "")

# 1. with SMC, prevalence settles at about 0.30
# 2. observed prevalence in 0-5 year olds during the program was ~0.28 —
#    a close match!
```

*Optional challenge: instead of the weighted-average shortcut, add
separate protected compartments ($P$ for protected-susceptible and
$I_P$ for protected-infected) to `sis_model`, with children moving into
protection when they receive SMC. Does it give the same answer?*

### Exercise 3.3: Interpreting the model

1.  Does the model without SMC reproduce the prevalence in the 6–10
    group? Does the model with SMC reproduce the prevalence in the 0–5
    group during the intervention? If the match is only approximate,
    what might explain the difference?
2.  This model treats each age group as a separate little world. In
    reality infected adults and older children can transmit to young
    children. Would accounting for that make the SMC effect on 0–5
    year olds look bigger or smaller than in your simple model?
3.  Our two surveys were only 12 months apart. Look at your simulation:
    how long does prevalence take to approach equilibrium after the
    force of infection changes? What does that imply about whether the
    during-survey prevalence in 0–5 year olds should be read as the
    *new equilibrium* or as a population still part-way through its
    decline?

## Extension: using the IgG data (optional)

IgG antibodies reflect **cumulative exposure**, not current infection,
so they cannot tell us about current prevalence directly. However, the
way antibody levels build up with age carries information about how
much transmission people have experienced over their lifetime.

Using `survey_igg`:

1.  Pick one antigen (e.g. AMA1) and plot IgG concentration against
    age, coloured by PCR status, faceted by survey. What do you notice?
2.  Define "seropositive" using a threshold (e.g. the 95th percentile
    of IgG in 0–5 year olds in the before survey) and compute the
    *proportion seropositive* by age group and survey. How does this
    age-seropositivity pattern differ from the age-prevalence pattern
    from Part 1, and why?
3.  Antibodies persist long after an infection is cleared, so IgG
    changes slowly. Compare seropositivity in 0–5 year olds between the
    two surveys, and compare that with the change in PCR prevalence you
    found in Exercise 2.4. Which moved more over the 12 months, and why
    does that make serology a *lagging* indicator of declining
    transmission?
4.  If transmission has genuinely dropped in recent years, which age
    groups would you expect to show the biggest difference between
    their antibody history and their current infection risk?

```{r}
# 1. IgG vs age, coloured by PCR status, faceted by survey
ama1 <- filter(survey_igg, antigen == "AMA1")
ggplot(ama1, aes(x = age, y = igg_conc, colour = factor(pcr_positive))) +
  geom_point(alpha = 0.4) +
  scale_y_log10() +
  facet_wrap(~survey) +
  labs(colour = "PCR positive", y = "IgG concentration (log scale)")
# IgG rises steadily with age (cumulative exposure) and is higher in
# PCR-positive individuals (recent infection boosts antibody levels)

# 2. seropositivity by age group and survey
threshold <- ama1 |>
  filter(survey == "before", age_group == "0-5") |>
  summarise(quantile(igg_conc, 0.95, na.rm = TRUE)) |>
  pull()

ama1 |>
  mutate(seropositive = igg_conc > threshold) |>
  group_by(survey, age_group) |>
  summarise(
    prop_seropositive = mean(seropositive, na.rm = TRUE),
    .groups = "drop"
  )
# seropositivity climbs steadily with age in both surveys — unlike PCR
# prevalence, it does not show a drop in 0-5 year olds during the program

# 3. PCR prevalence in 0-5 year olds fell substantially over 12 months,
#    but their seropositivity barely moved: antibodies from past
#    infections persist, so serology is a lagging indicator of
#    declining transmission.
```

## Summary

In this tutorial you have:

-   Cleaned and joined epidemiological datasets from two survey rounds,
    and handled missing data with complete-case analysis
-   Used chi-squared tests and multivariable linear models to show that
    SMC use, but not sex, ACT use, or bed net use, was associated with
    reduced malaria prevalence
-   Exploited the before/during design — including a
    difference-in-differences interaction model — to show that
    prevalence fell specifically in the SMC target group
-   Built an SIS compartmental model with `deSolve`, calibrated its
    force of infection from observed prevalence, and shown that an
    SMC-like intervention quantitatively reproduces the data
-   (Extension) Seen how serology complements prevalence data by
    capturing cumulative exposure, and why it responds more slowly to a
    drop in transmission
