---
title: "Workshop 6: Live-coding & Practice Exercises"
format:
  html:
    toc: true
    toc-location: left
    toc-title: "In this chapter:"
---

<!-- Companion file to workshop_6.qmd. -->
<!-- workshop_6.qmd is the complete, worked-example reference document. -->
<!-- This file contains: -->
<!--   1. "Live-coding" skeleton chunks for the instructor to build up on screen during class -->
<!--      (same variables/story as workshop_6.qmd, with key pieces blanked out as `___`) -->
<!--   2. "Your turn" practice questions for students to attempt themselves during class, -->
<!--      using different variables from `dosage` so they aren't just copying the demo -->
<!--   3. A collapsed instructor answer key at the very end -->

# Setup

Before starting, make sure the `dosage` data frame is loaded and cleaned exactly as in the main workshop:

```{r}
#| message: false
library(tidyverse)
library(tidyrstats)

dosage <- read_delim("data/mousezempic_dosage_data.csv") %>%
  mutate(
    mouse_strain = case_when(
      mouse_strain == "CD-1" ~ "CD1",
      mouse_strain == "Black 6" ~ "Black6",
      mouse_strain == "BALB C" ~ "BALBc",
      TRUE ~ mouse_strain
    )
  ) %>%
  na.omit(mouse_strain, sex)
```

# 1. Testing associations between categorical variables

## Live-coding (build together)

Follow along as we build the chi-squared test step by step. Fill in the blanks (`___`) as we go.

```{r}
#| eval: false

# Step 1: build a contingency table of mouse_strain and sex
strain_sex_table <- table(dosage$___, dosage$___)
strain_sex_table

# Step 2: visualise it as a stacked bar chart first
dosage %>%
  ggplot(aes(x = ___, fill = ___)) +
  geom_bar(position = "___") +
  labs(y = "Proportion", x = "Mouse strain", title = "Sex distribution by strain")

# Step 3: run the test
___(strain_sex_table)
```

## Your turn

Were mice allocated evenly across the three experimental **replicates** (`replicate`), regardless of `mouse_strain`? Investigate this using the same workflow we just built.

1. Build a contingency table of `mouse_strain` and `replicate`.
2. Visualise it as a stacked bar chart.
3. Check that the chi-squared assumption (expected cell counts) is reasonable before running the test.
4. Run the appropriate test and state your conclusion in plain language: is there evidence of an association between strain and replicate?

```{r}
#| eval: false

# Your code here

```

# 2. Correlation between continuous variables

## Live-coding (build together)

```{r}
#| eval: false

# Step 1: test the correlation
cor.test(dosage$___, dosage$___)

# Step 2: visualise it
ggplot(dosage, aes(x = ___, y = ___)) +
  geom_point() +
  geom_smooth(method = "___")

# Step 3: a correlation matrix across several variables at once
library(ggcorrplot)

dosage %>%
  select(drug_dose_g, weight_lost_g, tail_length_mm, initial_weight_g) %>%
  cor() %>%
  ggcorrplot(lab = ___)
```

## Your turn

Do heavier mice (at the start of the study) tend to lose more weight? And is there a relationship between a mouse's tail length and its starting weight?

1. Test the correlation between `initial_weight_g` and `weight_lost_g` using `cor.test()`.
2. Visualise this relationship with a scatterplot and a fitted line.
3. Test the correlation between `tail_length_mm` and `initial_weight_g`. Is the relationship positive or negative?
4. For one of the two relationships above, also compute the Spearman correlation. Does your conclusion change?

```{r}
#| eval: false

# Your code here

```

# 3. Parametric and non-parametric tests

## Live-coding (build together)

```{r}
#| eval: false

# Step 1: visualise the two groups
ggplot(dosage, aes(x = ___, y = weight_lost_g)) +
  geom_boxplot()

# Step 2: check normality with a QQ plot
ggplot(dosage, aes(sample = weight_lost_g)) +
  geom_qq() +
  geom_qq_line() +
  facet_wrap(~___)

# Step 3: confirm with a Shapiro-Wilk test (do this once per group)
dosage %>%
  filter(sex == "male") %>%
  pull(weight_lost_g) %>%
  ___()

# Step 4: run the t-test
dosage %>%
  t.test(weight_lost_g ~ ___, data = .)

# Step 5: run the non-parametric alternative
dosage %>%
  ___.test(weight_lost_g ~ sex, data = .)

# Step 6: for 3+ groups, use the Kruskal-Wallis test
dosage %>%
  kruskal.test(weight_lost_g ~ ___, data = .)
```

## Your turn

Does a mouse's starting weight (`initial_weight_g`) differ by sex? And does tail length (`tail_length_mm`) differ by strain?

1. Visualise `initial_weight_g` by `sex` with a boxplot, then check normality with a QQ plot and/or Shapiro-Wilk test for each group.
2. Based on your assumption check, run **either** a t-test **or** a Wilcoxon test (whichever is more appropriate) comparing `initial_weight_g` between sexes.
3. Visualise `tail_length_mm` by `mouse_strain` with a boxplot.
4. Run a Kruskal-Wallis test comparing `tail_length_mm` across the three strains, and describe what the boxplot and test result tell you together.

```{r}
#| eval: false

# Your code here

```

# 4. Linear modelling in the tidyverse

## Live-coding (build together)

```{r}
#| eval: false

# Step 1: a two-group comparison as a linear model
dosage %>%
  lm_test(weight_lost_g ~ ___)

# Compare the statistic/p.value above to the t-test from Section 3 — they should match!

# Step 2: group means tested against 0
dosage %>%
  lm_test(weight_lost_g ~ 0 + ___)

# Step 3: a continuous predictor
dosage %>%
  lm_test(weight_lost_g ~ ___)

# Step 4: check for a possible confounder before correcting for it
dosage %>%
  summarise(
    mean_dose = mean(drug_dose_g),
    mean_weight_lost = mean(weight_lost_g),
    .by = ___
  )

# Step 5: correct for the confounder
dosage %>%
  lm_test(weight_lost_g ~ drug_dose_g + ___)
```

## Your turn

Investigate whether `drug_dose_g` predicts `tail_length_mm`, and whether `mouse_strain` confounds that relationship. This one has a twist — pay close attention to the sign of the slope at each step!

1. Fit `lm_test(initial_weight_g ~ sex)`. Compare the `statistic` and `p.value` to your t-test from Section 3, Q2 — do they match?
2. Fit `lm_test(tail_length_mm ~ drug_dose_g)` (no covariates yet). Note the direction (sign) of the slope for `drug_dose_g`.
3. Compute the mean `drug_dose_g` and mean `tail_length_mm` for each `mouse_strain`. What do you notice about Black6 mice?
4. Now fit `lm_test(tail_length_mm ~ drug_dose_g + mouse_strain)`. What happens to the sign of the `drug_dose_g` slope compared to step 2? Discuss with your neighbour: why might correcting for a confounder actually **reverse** the direction of an effect, not just weaken it?

```{r}
#| eval: false

# Your code here

```

# 5. Logistic regression

## Live-coding (build together)

```{r}
#| eval: false

# Step 1: create a binary outcome
dosage <- dosage %>%
  drop_na(weight_lost_g, drug_dose_g) %>%
  mutate(
    responder = if_else(weight_lost_g > median(weight_lost_g), 1, 0),
    drug_dose_z = as.numeric(scale(___))
  )

dosage %>% count(___)

# Step 2: check for separation before fitting
table(dosage$mouse_strain, dosage$___)

# Step 3: fit the model
logit_model <- glm(
  responder ~ drug_dose_z + ___,
  data = dosage,
  family = "___"
)

summary(logit_model)

# Step 4: convert coefficients to odds ratios
broom::tidy(logit_model, exponentiate = ___, conf.int = TRUE)
```

## Your turn

Create a new binary outcome — whether a mouse has a **long tail** (above the sample median `tail_length_mm`) — and model it using `initial_weight_g` and `sex`.

1. Starting from `dosage`, create a new data frame `dosage_ex` with a `long_tail` column (1 if `tail_length_mm` is above the median, 0 otherwise), and a standardised `initial_weight_z` column (using `scale()`).
2. **Before fitting any model**, cross-tabulate `mouse_strain` against `long_tail`. What do you notice about one of the strains? Why would including `mouse_strain` as a predictor in a `glm()` here cause a problem?
3. Fit `glm(long_tail ~ initial_weight_z + sex, data = dosage_ex, family = "binomial")` instead.
4. Convert the coefficients to odds ratios with `broom::tidy(..., exponentiate = TRUE, conf.int = TRUE)`, and interpret both the `initial_weight_z` and `sex` odds ratios in plain language.

```{r}
#| eval: false

# Your code here

```

# Instructor answer key {.appendix}

::: {.callout-note title="Section 1 answer" collapse="true"}
```{r}
#| eval: false
strain_replicate_table <- table(dosage$mouse_strain, dosage$replicate)
strain_replicate_table

dosage %>%
  ggplot(aes(x = mouse_strain, fill = replicate)) +
  geom_bar(position = "fill")

chisq.test(strain_replicate_table)
```
Expected counts are comfortably above 5 in every cell, so the chi-squared approximation is valid. The p value is well above 0.05 (≈0.51), so there is no evidence of an association between strain and replicate — consistent with mice being allocated to replicates independently of strain.
:::

::: {.callout-note title="Section 2 answer" collapse="true"}
```{r}
#| eval: false
cor.test(dosage$initial_weight_g, dosage$weight_lost_g)
cor.test(dosage$tail_length_mm, dosage$initial_weight_g)
cor.test(dosage$tail_length_mm, dosage$initial_weight_g, method = "spearman")
```
`initial_weight_g` and `weight_lost_g` are moderately positively correlated (r ≈ 0.59, p < 0.001) — heavier mice tended to lose more weight. `tail_length_mm` and `initial_weight_g` are weakly negatively correlated (r ≈ -0.23, p < 0.001). Both conclusions hold under a Spearman correlation as well.
:::

::: {.callout-note title="Section 3 answer" collapse="true"}
```{r}
#| eval: false
shapiro.test(dosage$initial_weight_g[dosage$sex == "male"])
shapiro.test(dosage$initial_weight_g[dosage$sex == "female"])
# Both p < 0.05: normality is questionable in both groups
wilcox.test(initial_weight_g ~ sex, data = dosage)

kruskal.test(tail_length_mm ~ mouse_strain, data = dosage)
```
`initial_weight_g` is non-normal in both groups (Shapiro-Wilk p < 0.001), so a Wilcoxon test is the safer choice; it agrees with a t-test here (p < 0.001 either way) — males are heavier at baseline. Tail length differs strongly by strain (Kruskal-Wallis p < 0.001); the boxplot shows Black6 mice have noticeably **shorter** tails than BALBc or CD1.
:::

::: {.callout-note title="Section 4 answer" collapse="true"}
```{r}
#| eval: false
dosage %>% lm_test(initial_weight_g ~ sex)
dosage %>% lm_test(tail_length_mm ~ drug_dose_g)
dosage %>% summarise(mean_dose = mean(drug_dose_g), mean_tail = mean(tail_length_mm), .by = mouse_strain)
dosage %>% lm_test(tail_length_mm ~ drug_dose_g + mouse_strain)
```
The `lm_test(initial_weight_g ~ sex)` statistic and p value match the t-test exactly, confirming the equivalence.

This is a good example of **Simpson's paradox**: the uncorrected slope for `drug_dose_g` is strongly **negative** (higher dose associated with shorter tails), but Black6 mice happen to have both a higher average dose *and* much shorter tails than the other strains. Once `mouse_strain` is added to the model, the `drug_dose_g` slope flips to strongly **positive** — within any given strain, higher doses are actually associated with longer tails. The uncorrected relationship was driven entirely by strain differences, not a true dose effect in the opposite direction. This is worth discussing explicitly: correcting for a confounder can change not just the strength but the *direction* of an estimated effect.
:::

::: {.callout-note title="Section 5 answer" collapse="true"}
```{r}
#| eval: false
dosage_ex <- dosage %>%
  drop_na(tail_length_mm, initial_weight_g) %>%
  mutate(
    long_tail = if_else(tail_length_mm > median(tail_length_mm), 1, 0),
    initial_weight_z = as.numeric(scale(initial_weight_g))
  )

table(dosage_ex$mouse_strain, dosage_ex$long_tail)

ex_logit <- glm(long_tail ~ initial_weight_z + sex, data = dosage_ex, family = "binomial")
broom::tidy(ex_logit, exponentiate = TRUE, conf.int = TRUE)
```

The cross-tab reveals **perfect separation**: every single Black6 mouse falls in the short-tail group (0 long-tailed Black6 mice out of 119). Including `mouse_strain` as a predictor of `long_tail` would make the model unstable (infinite/unreliable coefficient for Black6), so it must be left out of this particular model.

Both `initial_weight_z` and `sex` are significant predictors: each 1-SD increase in starting weight is associated with roughly a 59% *decrease* in the odds of having a long tail (OR ≈ 0.41), while male mice have roughly 6-fold higher odds of a long tail than female mice (OR ≈ 6.2), holding starting weight constant.
:::
